Reformulating phylogenetic mixed models to improve flexibility and speed

### 2022-08-09 phyloglmm_ms_resubmit.tex
cleanup, tweaks

% general-purpose mathrm macros
\newcommand{\symsub}[2]{\ensuremath{#1_{\tiny \mathrm{#2}}}\xspace}
\newcommand{\mrm}[1]{\ensuremath{\mathrm{#1}}
\xspace}

@@

\newcommand{\plmefour}{phyloglmm/lme4}
\newcommand{\pgTMB}{phyloglmm/glmmTMB}

% \pagenumbering{gobble}
\linenumbers

@@

In classic PRs, phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the response trait evolves along the branches of the phylogeny according to a Brownian-motion (BM) evolutionary model \citep{felsenstein1985phylogenies}. 
If the residuals are normally distributed and observed without additional error or within-species variation, Felsenstein's method of phylogenetically independent contrasts  \citep[PICS:][]{felsenstein1985phylogenies} is sufficient to account for the phylogenetic correlation.
More recent approaches --- including phylogenetic generalized linear mixed models  \citep[PGLMM:][]{ives2011generalized, housworth2004phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- extend PICs by considering different (non-Gaussian) response distributions and by accounting for evolutionary models other than BM.
These methods partition residual variation into two components: (1) independent residual variation (tip variation, which is confounded with observation error) and (2) phylogenetic signal (evolutionary process error) \citep{hansen2012interpreting, housworth2004phylogenetic}.
If each species' traits are measured multiple times, we can distinguish a third level of variation; in this case, phylogenetic variation and tip variation are both part of the evolutionary process error while the observation error can be independently identified from within-species variation \citep{kostikova2016bridging, devillemereuil2012bayesian}.

@@

Several recent studies developed new tools to fit phylogenetic mixed models in
community ecology applications \citep{nowakowski2018phylogenetic, li2017canfun}.
However, the phylogenetic mixed modeling tools that allow extensions such as random slopes
and separation of tip and observation error in a frequentist framework are inflexible; thus, biologists needing to fit more complex models typically turn to Bayesian approaches which is the state of the art, despite their additional computational burden \citep{hadfield2010mcmc, burkner2018brms, kostikova2016bridging} (Table \ref{table:model}).

\begin{table}[]
\begin{tabular}{|l|l|l|l|}
\hline
\textbf{Model} & \textbf{Method} & \textbf{Data} & \textbf{Platform} \\ \hline
\multirow{2}{*}{
\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Model (GLM)
\end{tabular}} & Correlated residual & \begin{tabular}[c]{@{}l@{}}Single observation \\ per species \end{tabular} & \begin{tabular}[c]{@{}l@{}}\pkg{nlme}:gls, \\ \pkg{ape}:pic\end{tabular} \\ \cline{2-4} 
& \begin{tabular}[c]{@{}l@{}}Residual \\ + phylogenetic intercept\end{tabular} & \begin{tabular}[c]{@{}l@{}}Single observation \\ per species \end{tabular} & \begin{tabular}[c]{@{}l@{}}Pagel's $\lambda$\\ Blomberg's $k$ \\ via \pkg{nlme}:gls\\ \pkg{phylolm}\end{tabular} \\ \hline
\multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Mixed Model (GLMM)\end{tabular}} & \multirow{2}{*}{Random effect}                                        & \begin{tabular}[c]{@{}l@{}}Single observation \\per species, \\ Balanced design\end{tabular} & \pkg{pez}                                                                                            \\ \cline{3-4} 
                                                                                                  &                                                                       & Unrestricted                                                                 & \pkg{\plmefour}, \pkg{\pgTMB}, \pkg{phyr}                                                                                  \\ \hline
\multirow{2}{*}{Bayesian GLMM}                                                                    & \multirow{2}{*}{Random effect}                                        & Balanced design                                                              & \pkg{MCMCglmm}                                                                                       \\ \cline{3-4} 
                                                                                                  &                                                                       & Unrestricted                                                                 & \pkg{brms}                                                                                           \\ \hline
\end{tabular}
\caption{GLM-based phylogenetic comparative methods and their implementations in R packages.}
\label{table:model}
\end{table}

@@

The phylogenetic generalized linear mixed model (PGLMM) framework defines a wider range of models that includes the standard phylogenetic regression as a special case \citep{lynch1991methods}.
The PGLMM allows for non-Gaussian responses and uses incorporates multiple components of variability.
The PGLMM has the form:
\newcommand{\dist}{{\cal D}}
\begin{equation}
\begin{aligned}
\bY & = \dist(g^{-1}(\bmu), \phi) \\
\bmu & = \bX \bbeta + \bZ \bb  \\
\bb & \sim \textrm{MVN}(0, \bSigma(\theta))  \\
\label{eq:glmm}
\end{aligned}
\end{equation}
where in addition to the terms from (\ref{eq:gls}) $\bZ$ is an $n \times q$ model matrix for the $q$-dimensional vector-valued random effects; $\bb$ represents the conditional mean (or mode) of the random effect, which is multivariate normally distributed with covariance matrix $\bSigma(\theta)$; and $\phi$ is a scale parameter for the conditional distribution $\dist$.
When $\dist$ is Gaussian, $g$ is the identity function, $\bZ$ is the identity matrix, and $\bSigma(\theta)=\sigma^2\bC$, (\ref{eq:glmm}) reduces to (\ref{eq:gls}).

@@

In general, correlations within statistical models can be integrated either in the 
covariance matrix $\bSigma$ or in the structure of the model matrix $\bZ$ \citep{hefley2017basis};
thus, we can use $\bZ$ to incorporate phylogenetic correlations.
Suppose evolution follows a BM process, i.e., continuous traits evolve independently, following an unbiased random walk along each branch of the phylogeny.
Then the phylogenetic variability of a particular species can be written as the sum of the variances of evolutionary changes that occurred on all of the branches in its history. 
Thus, modeling the evolutionary history of each species with a sequence of independent errors with species--branch matrix $\bS$ is equivalent to imposing a correlation $\bC$.
For example, for the phylogeny in figure \ref{fig:tree}, the corresponding $\bS$ takes the form:

\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.8]{./figure/phylotree.png}
  \caption{Three-species phylogenetic tree.}
\label{fig:tree}
\end{figure}
\end{center}
\[
  \begin{blockarray}{ccccc}
  & L_1 & L_2 & L_3 & L_4  \\
  \begin{block}{c(cccc)}
  t_1 & \ell_1 & \ell_2 & 0           & 0 \\
  t_2 & \ell_1 &  0          & \ell_3 & 0 \\
  t_3 & 0           &  0          & 0           & \ell_4 \\
  \end{block}
  \end{blockarray}
  \hspace{0.5cm}
  \begin{blockarray}{c}
    \vphantom{L_1} \\
    \begin{block}{(c)}
      \e_1 \\
      \e_2 \\
      \e_3 \\
      \e_4 \\
    \end{block}
  \end{blockarray}
  \]
% \begin{pmatrix}
% \e_1 \\
% \e_2 \\
% \e_3 \\
% \e_4
% \end{pmatrix}
%% FIXME: double check this
For example, the phylogenetic effect for species 1 is $\ell_1 \e_1 + \ell_2 \e_2$, where $\ell_i = \sqrt{L_i}$, the square root of the branch length $L_i$ in Figure \ref{fig:tree}, and the $\e_i$ are independent Normal deviates with zero mean and variance $\sigma^2$ (i.e. the phylogenetic variance for species 1 is $\textrm{E}[(\ell_1 \e_1 + \ell_2 \e_2)^2] = (L_1 + L_2)\sigma^2$).
% The variance of Brownian-motion, is the evolutionary rate, where it is linear in evolution time.

@@

\[
\bS_{ind} = \begin{bmatrix}
1 & 1 & 0 & 0 \\ 
1 & 0 & 1 & 0 \\ 
0 & 0 & 0 & 1
\end{bmatrix} , 
\qquad
\bell = \begin{bmatrix}
\ell_1 \\
\ell_2 \\
\ell_3 \\
\ell_4 
\end{bmatrix} .
\]
$\bS_{ind}$ is a binary (indicator) matrix that describes whether a particular branch occurs in the history of a focal species. 
$\bS \bS^T$ gives the covariance matrix of the phylogeny.

@@

\begin{equation}
\begin{aligned}
\bZ^{C}_{i} = (\bS^{\top}\bJ^{\top}_{i} \ast \bX^{\top}_{i})^{\top} =
\left[
\left(
\begin{bmatrix}
\ell_{1} & \ell_{1}  & 0 \\
\ell_{2} &  0  & 0 \\
0  &  \ell_{3} & 0 \\
0 & 0 &  \ell_{4} 
\end{bmatrix}
\begin{bmatrix}
1 & 0  & 0 \\
0 & 1  & 0 \\
0 & 0  & 1  
\end{bmatrix}
\right)
\ast
\begin{bmatrix}
1   & 1   & 1  \\ 
t_1 & t_2 & t_3
\end{bmatrix} 
\right]^{\top}
\\
= \begin{bmatrix}
\ell_{1} & \ell_{1}t_1 & \ell_{2} & \ell_{2}t_1 & 0 & 0 & 0 & 0 \\
\ell_{1} & \ell_{1}t_2 & 0 & 0 & \ell_{3} & \ell_{3}t_2 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 0 & \ell_{4} & \ell_{4}t_3
\end{bmatrix}.
\end{aligned}
\end{equation}

@@

We generated test data based on (\ref{eq:glmm}) with a single response variable $\bY$ and a single normally distributed predictor variable $\bm{t}$ for $n$ = 25, 50, and 100 species.
The response variable $\bY$ is conditionally normally distributed (i.e., $\cal{D}$ is a Gaussian distribution, and $g$ is the identity link function), corresponding to a linear mixed effect model. 
For the first set of simulations, we simulate one observation per species.
% For a broader range of comparisons, we simulate one observation ($X_i,Y_i$) per species.
Thus, the full simulation model is as follows:
\begin{equation}
\begin{aligned}
\bY & = \dist(g^{-1}(\bmu), \phi) \\
\bmu & = (\bf{\beta_0 + b_{\mathrm{phy_{int}}}}) + (\bf{\beta_1 + b_{\mathrm{phy_{slope}}}}) \bm{t} + \besp \\\
(\bf{b_{\mathrm{phy_{int}}}, b_{\mathrm{phy_{slope}}}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
\Sigma^2_{\mathrm{phy_{int}}} & \Sigma_{\mathrm{phy_{int-slope}}} \\ 
\Sigma_{\mathrm{phy_{int-slope}}} & \Sigma^2_{\mathrm{phy_{slope}}}
\end{bmatrix} 
\right) \\ 
\besp & \sim \textrm{N} ( 0 , \sigma_{\epsilon}^2) \quad.
\label{eq:single_glmm}
\end{aligned}
\end{equation}
The model has two fixed effect parameters ($\beta_0$ and $\beta_1$), three random effect parameters (phylogenetic random intercept variance $\Sigma^2_{\mathrm{phy}_{int}}$, phylogenetic random slope variance $\Sigma^2_{\mathrm{phy}_{slope}}$ and covariance between phylogenetic random intercept and slope $\Sigma_{\mathrm{phy_{int-slope}}}$) and residual variance ($\sigma_{\epsilon}^2$).  
The covariance between phylogenetic random intercept and slope measures the correlation of phylogenetic effects on the slope ($b_{\mathrm{phy_{slope}}}$) and intercept ($b_{\mathrm{phy_{int}}}$) for each branch of the phylogeny; i.e. a positive correlation indicates that species with similar intercepts also have similar slopes.
Predictor-level and intercept-level random effects of species are not applicable in this simulation setting because there is only a single observation per species, so within-species variation cannot be separated from tip variation.

@@

We extend the simulation model by adding multiple groups where each group has one observation per species. 
The multi-group model is a generalization of multiple-site models used in community ecology to model phylogenetic attraction \citep{helmus2007separating}. 
The full multi-group model is as follows: 
\begin{equation}
\begin{aligned}
\bY & = \dist(g^{-1}(\bmu), \phi) \\
\bmu & = (\bf{\beta_0 + b_{\mathrm{phy_{int}}} + b_{\mathrm{sp_{int}}} + b_{\mathrm{group}}}) + (\bf{\beta_1 + b_{\mathrm{phy_{slope}}} + b_{\mathrm{sp_{slope}}}}) \bm{t} + \bf{b_{\mathrm{sp:group}}} + \besp \\
(\bf{b_{\mathrm{phy_{int}}}, b_{\mathrm{phy_{slope}}}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
\sigma^2_{\mathrm{phy_{int}}} & \sigma_{\mathrm{phy_{int-slope}}} \\ 
\sigma_{\mathrm{phy_{int-slope}}} & \sigma^2_{\mathrm{phy_{slope}}}
\end{bmatrix}
\right) \\
(\bf{b_{\mathrm{sp_{int}}}, b_{\mathrm{sp_{slope}}}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
\sigma^2_{\mathrm{sp_{int}}} & \sigma_{\mathrm{sp_{int-slope}}} \\ 
\sigma_{\mathrm{sp_{int-slope}}} & \sigma^2_{\mathrm{sp_{slope}}}
\end{bmatrix}
\right) \\
\bf{b_{\mathrm{group}}} & \sim \textrm{MVN} ( 0 , \sigma^2_{\mathrm{group}}) \\
\bf{b_{\mathrm{sp:group}}} & \sim \textrm{MVN} (0, \mathrm{I}_{\mathrm{group}} \otimes \sigma^2_{phy}) \\
\besp & \sim \textrm{N} ( 0 , \sigma_{\epsilon}^2),
\label{eq:multiple_glmm}
\end{aligned}
\end{equation}
where $\textrm{I}_{\textrm{group}}$ is a indicator matrix for groups and $\otimes$ is the Kronecker product.

Compared to the single--group model, the multi-group simulation model has five additional random effect parameters (predictor-level ($\sigma^2_{\mathrm{sp_{slope}}}$) and intercept-level ($\sigma^2_{\mathrm{sp_{int}}}$) random effect of species variance and their covariance ($\sigma_{\mathrm{sp_{int-slope}}}$), random intercept of group ($\bf{b_{\mathrm{group}}}$) and random intercept of species-group interaction ($\bf{b_{\mathrm{sp:group}}}$)).
Because each species has multiple observations we can distinguish variation among species from residual variation, and thus we can include predictor-level and intercept-level random effects of species.
Variance in the intercept of species--group interactions ($\bf{b_{\mathrm{sp:group}}}$) describes whether the species within a group have more similar responses on average than expected by chance, equivalent to phylogenetic attraction \citep{helmus2007separating}.

@@

We compare our approach with five other R packages that can fit PRs: \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez}, \pkg{phyr} \citep{phyr} and \pkg{brms} \citep{burkner2018brms}.
Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used PR models; it fits a linear model with a covariance structure that assumes an evolutionary process on the tree (typically BM, but other processes can be used) instead of treating the residual error for each species as independent.
Phylogenetic generalized linear models (PGLM) (\code{phyloglm} in the \pkg{phylolm} package) are extend PGLS by allowing for both phylogenetic and residual variation, as well as non-Gaussian response variables.
Both \code{gls} and \code{phylolm} can model non-Brownian evolutionary processes and different correlation structures (e.g., Pagel's $\lambda$ or Blomberg's $K$), but we restrict our PGLS fits to the simple BM correlation.
Neither PGLS nor PGLM can handle random slopes or multiple observations within a species.
Among the few packages that currently fit phylogenetic slopes to predictor level variation are \pkg{pez} and \pkg{phyr}, which can handle random-slope models ($\sigma^2_{\mathrm{phy_{slope}}}$) and random intercepts of species-group interactions ($\bf{b_{\mathrm{sp:group}}}$) but do not incorporate covariation between phylogenetic random slopes and intercepts ($\sigma_{\mathrm{phy_{int-slope}}}$).
Lastly, Bayesian PGLMMs using Markov chain Monte Carlo (MCMC) can handle all of the cases described above. 
However, MCMC is usually much more computationally expensive for GLMMs than platforms using deterministic optimization.
\pkg{MCMCglmm} \citep{hadfield2010general} is the most widely used Bayesian phylogenetic GLMM, and recently \pkg{brms}, which uses a more computationally efficient MCMC technique.

@@

Using the R package \pkg{ape} \citep{ape}, we simulated 100 random phylogenetic trees  for each sample size (n = 25, 50, 100 and an additional n = 500 for the multi-group model) and then simulated the responses for each tree (\ref{eq:single_glmm}, \ref{eq:multiple_glmm}). 
Each realization was fitted using all model variants. 
All simulation parameters are shown in Figure \ref{ssplot} and Figure \ref{msplot}. 
Table \ref{table:platform} shows the parameters that are estimable for each platform. 
We only evaluated the goodness of fit for model fits that passed the convergence tests implemented by the package.
For Bayesian fits, we evaluate realizations with Gelman-Rubin statistic $<1.1$. Based on recent concerns about Gelman-Rubin thresholds \citep{vats2018revisiting}), we additionally restricted results to fits with  effective sample size $>1000$ for the fixed effect parameters ($\beta_{0}$ and $\beta_{1}$) \citep{vehtari2019rank}. 
For each replicate, we sample two chains starting with 10000 iterations. 
We first evaluate our estimates by looking at the distribution of the estimated values (maximum likelihood estimates for non-Bayesian platforms and posterior medians for Bayesian platforms) to quantify bias and variance (i.e., quality of the point estimate).
We computed 95\% Wald confidence intervals for frequentist methods and quantile-based intervals for Bayesian methods, then computed coverage --- the proportion of simulations in which the computed confidence intervals include the true values of parameters --- to assess the quality of the confidence intervals.
We also compare computational speed between different platforms to evaluate the efficiency of different platforms and methods.

>>>>

% general-purpose mathrm macros
\newcommand{\symsub}[2]{\ensuremath{#1_{\tiny \mathrm{#2}}}\xspace}
\newcommand{\mrm}[1]{\ensuremath{\mathrm{#1}}\xspace}

@@

\newcommand{\plmefour}{phyloglmm/lme4}
\newcommand{\pgTMB}{phyloglmm/glmmTMB}
\newcommand{\phyvar}{\sigma} %% phylogenetic var: allow sigma/Sigma switch
\newcommand{\phyv}[1]{\ensuremath{\phyvar^2_{\mathrm{phy}_\mathrm{#1}}}}
\newcommand{\phys}[1]{\ensuremath{\phyvar_{\mathrm{phy}_\mathrm{#1}}}}
\newcommand{\bphys}[1]{\ensuremath{{\bf b}_{\mathrm{phy}_{\mathrm{#1}}}}}
\newcommand{\bsps}[1]{\ensuremath{{\bf b}_{\mathrm{sp}_{\mathrm{#1}}}}}
% \pagenumbering{gobble}
\linenumbers

@@

In classic PRs, phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the response trait evolves along the branches of the phylogeny according to a Brownian-motion (BM) evolutionary model \citep{felsenstein1985phylogenies}. 
If the residuals are normally distributed and observed without additional error or within-species variation, Felsenstein's method of phylogenetically independent contrasts  \citep[PICS:][]{felsenstein1985phylogenies} is sufficient to account for the phylogenetic correlation.
More recent approaches --- including phylogenetic generalized linear mixed models  \citep[PGLMM:][]{ives2011generalized, housworth2004phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- extend PICs by considering different (non-Gaussian) response distributions and by accounting for evolutionary models other than BM.
These methods partition residual variation into two components: (1) independent residual variation (tip variation, which may be confounded with observation error) and (2) phylogenetic signal  \citep[evolutionary process error:][]{hansen2012interpreting, housworth2004phylogenetic}.
If each species' traits are measured multiple times, we can distinguish a third level of variation; in this case, phylogenetic variation and tip variation both contribute to the evolutionary variation while the observation error can be independently identified from within-species variation \citep{kostikova2016bridging, devillemereuil2012bayesian}.

@@

Several recent studies have developed new tools to fit phylogenetic mixed models in
community ecology applications \citep{nowakowski2018phylogenetic, li2017canfun}.
However, the phylogenetic mixed modeling tools that allow extensions such as random slopes
and separation of tip and observation error in a frequentist framework may not allow other extensions like non-normal response distributions or additional random effects; thus, biologists needing to fit more complex models typically turn to more flexible Bayesian approaches, despite their additional computational burden \citep{hadfield2010mcmc, burkner2018brms, kostikova2016bridging} (Table \ref{table:model}).

\input table1.tex

@@

The phylogenetic generalized linear mixed model (PGLMM) framework defines a wider range of models that includes the standard phylogenetic regression as a special case \citep{lynch1991methods}.
The PGLMM allows for non-Gaussian responses and incorporates multiple components of variability.
The PGLMM has the form:
\newcommand{\dist}{{\cal D}}
\begin{equation}
\begin{aligned}
\bY & \sim \dist(\bmu, \phi) \\
\bmu & = g^{-1}(\bX \bbeta + \bZ \bb)  \\
\bb & \sim \textrm{MVN}(0, \bSigma(\theta))  \\
\label{eq:glmm}
\end{aligned}
\end{equation}
where in addition to the terms from (\ref{eq:gls}) $\bZ$ is an $n \times q$ model matrix for the $q$-dimensional vector-valued random effects; $\bb$ represents the conditional mean (or mode) of the random effect, which is multivariate normally distributed with covariance matrix $\bSigma(\theta)$; and $\phi$ is a scale parameter for the conditional distribution $\dist$.
The PGLMM reduces to the simple PR model (\ref{eq:gls}) when
$\dist$ is Gaussian, $g$ is the identity function, $\bZ$ is the identity matrix,
$\bSigma(\theta)=\sigma^2\bC$, and $\dist$ is Gaussian. In addition, we need $\phi=\sigma_r^2 = 0$ so that the residual variance disappears and the only variance comes from the phylogenetic covariance matrix; otherwise, the additional residual variance term corresponds to one implementation of Pagel's $\lambda$.

@@

In general, correlations within statistical models can be integrated either in the 
covariance matrix $\bSigma$ or in the structure of the model matrix $\bZ$ \citep{hefley2017basis};
thus, we can use $\bZ$ to incorporate phylogenetic correlations.
Suppose evolution follows a BM process, i.e., continuous traits evolve independently at a constant rate, following an unbiased random walk along each branch of the phylogeny.
Then the phylogenetic variability of a particular species can be written as the sum of the variances of evolutionary changes that occurred on all of the branches in its history. 
Thus, modeling the evolutionary history of each species with a sequence of independent errors with species--branch matrix $\bS$ is equivalent to imposing a correlation $\bC$.
For example, for the phylogeny in figure \ref{fig:tree}, the corresponding $\bS$ takes the form:

\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.8]{./figure/phylotree.png}
  \caption{Three-species phylogenetic tree.}
\label{fig:tree}
\end{figure}
\end{center}

\input matmult.tex

% \begin{pmatrix}
% \e_1 \\
% \e_2 \\
% \e_3 \\
% \e_4
% \end{pmatrix}
%% FIXME: double check this
For example, the phylogenetic effect for species 1 is $\ell_1 \e_1 + \ell_2 \e_2$, where $\ell_i = \sqrt{L_i}$, the square root of the branch length $L_i$ in Figure \ref{fig:tree}, and the $\e_i$ are independent Normal deviates with zero mean and variance $\sigma^2$ (i.e. the phylogenetic variance for species 1 is $\textrm{E}[(\ell_1 \e_1 + \ell_2 \e_2)^2] = (L_1 + L_2)\sigma^2$).
% The variance of Brownian-motion, is the evolutionary rate, where it is linear in evolution time.

@@

\[
\bS_{ind} = \begin{bmatrix}
1 & 1 & 0 & 0 \\ 
1 & 0 & 1 & 0 \\ 
0 & 0 & 0 & 1
\end{bmatrix} , 
\qquad
\bell = \begin{bmatrix}
\ell_1 \\
\ell_2 \\
\ell_3 \\
\ell_4 
\end{bmatrix} .
\]
$\bS_{ind}$ is a binary matrix that describes whether a particular branch occurs in the history of a focal species. 
$\bS \bS^T$ gives the covariance matrix of the phylogeny.

@@

\begin{equation}
\begin{aligned}
\bZ^{C}_{i} = (\bS^{\top}\bJ^{\top}_{i} \ast \bX^{\top}_{i})^{\top} =
\left[
\left(
\begin{bmatrix}
\ell_{1} & \ell_{1}  & 0 \\
\ell_{2} &  0  & 0 \\
0  &  \ell_{3} & 0 \\
0 & 0 &  \ell_{4} 
\end{bmatrix}
\begin{bmatrix}
1 & 0  & 0 \\
0 & 1  & 0 \\
0 & 0  & 1  
\end{bmatrix}
\right)
\ast
\begin{bmatrix}
1   & 1   & 1  \\ 
t_1 & t_2 & t_3
\end{bmatrix} 
\right]^{\top}
\\
= \begin{bmatrix}
\ell_{1} & \ell_{1}t_1 & \ell_{2} & \ell_{2}t_1 & 0 & 0 & 0 & 0 \\
\ell_{1} & \ell_{1}t_2 & 0 & 0 & \ell_{3} & \ell_{3}t_2 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 0 & \ell_{4} & \ell_{4}t_3
\end{bmatrix}.
\end{aligned}
\label{eq:zmatrix}
\end{equation}

@@

Using the formulation described in (\ref{eq:glmm}--\ref{eq:zmatrix}), we generated test data with a single response variable $\bY$ and a single normally distributed predictor variable $\bm{t}$ for $n$ = 25, 50, and 100 species.
The response variable $\bY$ is conditionally normally distributed (i.e., $\cal{D}$ is a Gaussian distribution, and $g$ is the identity link function), corresponding to a linear mixed effect model. 
For the first set of simulations, we simulate one observation per species.
% For a broader range of comparisons, we simulate one observation ($X_i,Y_i$) per species.
Thus, the full simulation model is as follows:
\begin{equation}
\begin{aligned}
\bY & = \dist(g^{-1}(\bmu), \phi) \\
\bmu & = (\beta_0 + \bphys{int}) + (\beta_1 + \bphys{slope}) \bm{t} + \besp \\\
(\bphys{int}, \bphys{slope}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
\phyv{int} & \phys{int-slope} \\ 
\phys{int-slope} & \phyv{slope}
\end{bmatrix} 
\right) \\ 
\besp & \sim \textrm{N} ( 0 , \sigma_{\epsilon}^2) \quad.
\label{eq:single_glmm}
\end{aligned}
\end{equation}
The model has two fixed effect parameters ($\beta_0$ and $\beta_1$), three random effect parameters (phylogenetic random intercept variance (\phyv{int}), phylogenetic random slope variance (\phyv{slope}) and covariance between phylogenetic random intercept and slope (\phys{int-slope}) and residual variance ($\sigma_{\epsilon}^2$).  
The covariance between phylogenetic random intercept and slope measures the correlation of phylogenetic effects on the slope ($b_{\mathrm{phy_{slope}}}$) and intercept ($b_{\mathrm{phy_{int}}}$) for each branch of the phylogeny; i.e. a positive correlation indicates that species with similar intercepts also have similar slopes.
Predictor-level and intercept-level random effects of species are not applicable in this simulation setting because there is only a single observation per species, so within-species variation cannot be separated from tip variation.

@@

We extend the simulation model by adding multiple groups where each group has one observation per species. 
The multi-group model is a generalization of multiple-site models used in community ecology to model phylogenetic attraction and repulsion \citep{helmus2007separating}. 
The full multi-group model is as follows: 
\begin{equation}
\begin{aligned}
\bY & = \dist(g^{-1}(\bmu), \phi) \\
\bmu & = (\beta_0 + \bphys{int} + \bsps{int} + {\bf b}_{\mathrm{group}}}) + (\beta_1 + \bphys{slope} + \bsps{slope}) \bm{t} + {\bf b}_{\mathrm{sp:group}}} + \besp \\
(\bphys{int}, \bphys{slope}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
\phyv{int} & \phys{int-slope} \\ 
\phys{int-slope} & \phyv{slope}
\end{bmatrix}
\right) \\
(\bf{b_{\mathrm{sp_{int}}}, b_{\mathrm{sp_{slope}}}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
\sigma^2_{\mathrm{sp_{int}}} & \sigma_{\mathrm{sp_{int-slope}}} \\ 
\sigma_{\mathrm{sp_{int-slope}}} & \sigma^2_{\mathrm{sp_{slope}}}
\end{bmatrix}
\right) \\
\bf{b_{\mathrm{group}}} & \sim \textrm{MVN} ( 0 , \sigma^2_{\mathrm{group}}) \\
\bf{b_{\mathrm{sp:group}}} & \sim \textrm{MVN} (0, \mathrm{I}_{\mathrm{group}} \otimes \sigma^2_{\mathrm{phy}}) \\
\besp & \sim \textrm{N} ( 0 , \sigma_{\epsilon}^2),
\label{eq:multiple_glmm}
\end{aligned}
\end{equation}
where $\textrm{I}_{\textrm{group}}$ is a indicator matrix for groups and $\otimes$ is the Kronecker product.

\bmb{ML: please check notation; in particular, is $\sigma^2_{\mathrm{phy}}$ correct??}
Compared to the single--group model, the multi-group simulation model has five additional random effect parameters: predictor-level ($\sigma^2_{\mathrm{sp_{slope}}}$) and intercept-level ($\sigma^2_{\mathrm{sp_{int}}}$) among-species variances;  their covariance ($\sigma_{\mathrm{sp_{int-slope}}}$);  among-group variation in the intercept ($\sigma^2_{\mathrm{group}}$); and variation among species-group combinations in the intercept ($\sigma^2_{\mathrm{phy}}$).
Because each species has multiple observations we can distinguish variation among species from residual variation, and thus we can include predictor-level and intercept-level random effects of species.
Variance in the intercept of species--group interactions ($\sigma^2_{\mathrm{phy}}$) describes whether the species within a group have more similar responses on average than expected by chance, equivalent to phylogenetic attraction \citep{helmus2007separating}.

@@

We compare our approach with five other R packages that can fit PRs: \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez}, \pkg{phyr} \citep{phyr} and \pkg{brms} \citep{burkner2018brms}.
Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used PR models; it fits a linear model with a covariance structure that assumes an evolutionary process on the tree (typically BM, but other processes can be used) instead of treating the residual error for each species as independent.
Phylogenetic generalized linear models (PGLM) (\code{phyloglm} in the \pkg{phylolm} package) extend PGLS by allowing for both phylogenetic and residual variation, as well as non-Gaussian response variables.
Both \code{gls} and \code{phylolm} can model non-Brownian evolutionary processes and different correlation structures (e.g., Pagel's $\lambda$ or Blomberg's $K$), but we restrict our PGLS fits to the simple BM correlation.
Neither PGLS nor PGLM can handle random slopes or multiple observations within a species.
Among the few packages that currently fit phylogenetic slopes to predictor level variation are \pkg{pez} and \pkg{phyr}, which can handle random-slope models ($\sigma^2_{\mathrm{phy_{slope}}}$) and random intercepts of species-group interactions ($\bf{b_{\mathrm{sp:group}}}$) but do not incorporate covariation between phylogenetic random slopes and intercepts ($\sigma_{\mathrm{phy_{int-slope}}}$).
Lastly, Bayesian PGLMMs using Markov chain Monte Carlo (MCMC) can handle all of the cases described above. 
However, MCMC is usually much more computationally expensive for GLMMs than platforms using deterministic optimization.
\pkg{MCMCglmm} \citep{hadfield2010general} is the most widely used Bayesian phylogenetic GLMM;  the more recent\pkg{brms} package is extremely flexible and uses Hamilton Monte Carlo, which is often more computationally efficient (although below we find that \pkg{MCMCglmm} is faster in this application).

@@

Using the R package \pkg{ape} \citep{ape}, we simulated 100 random phylogenetic trees  for each sample size ($n$ = 25, 50, 100 and an additional $n$ = 500 for the multi-group model) and then simulated the responses for each tree (\ref{eq:single_glmm}, \ref{eq:multiple_glmm}). 
Each realization was fitted using all model variants. 
All simulation parameters are shown in Figure \ref{ssplot} and Figure \ref{msplot}. 
Table \ref{table:platform} shows the parameters that are estimable for each platform. 
We only evaluated the goodness of fit for model fits that passed the convergence tests implemented by the package.
For Bayesian fits, we evaluate realizations with Gelman-Rubin statistic $<1.1$. Based on recent concerns about Gelman-Rubin thresholds \citep{vats2018revisiting}, we additionally restricted results to fits with  effective sample size $>1000$ for the fixed effect parameters  \citep[$\beta_{0}$ and $\beta_{1}$: ][]{vehtari2019rank}. 
For each replicate, we sample two chains starting with 10000 iterations. 
We first evaluate our estimates by looking at the distribution of the estimated values (maximum likelihood estimates for non-Bayesian platforms and posterior medians for Bayesian platforms) to quantify bias and variance (i.e., quality of the point estimate).
We computed 95\% Wald confidence intervals for frequentist methods \bmb{check: I would think lme4 would only give profile CIs for RE parameters?} and quantile-based intervals for Bayesian methods, then computed coverage --- the proportion of simulations in which the computed confidence intervals include the true values of parameters --- to assess the quality of the confidence intervals.
We also compare computational speed between different platforms.
